They're Clutching up! Team Momentum in Round-Based Esports

### 2022-08-21 paper/paper.tex
re-reading, added all tables, by p-value adj, no p-values in tables

\begin{abstract}
My research investigates patterns in round win percentages in
professional Search and Destroy (SnD) matches of the popular
first-person shooter game Call of Duty (CoD). First, I find evidence
supprting the hypothesis that round win probability can be modeled as a
constant across the series, although not at the naive 50\%. Second, I
examine post-streak round win probability, given the series outcome. I
find no evidence that post-streak win rate is significantly different
than expected for combinations of streak length, series length, and
series winner, suggesting that the perception of momentum at the team
level is deceiving.
\end{abstract}

@@

I adopt the terminology ``series'' to refer to what CoD SnD players
typically call a ``match'', so as to emulate the terminology of playoff
series in professional leagues like the National Basketball Association,
National Hockey League, and Major League Baseball. A ``game'' or a
``match'' in such leagues is analogous to a ``round'' of CoD SnD.

@@

The data set consists of all SnD matches played in major tournaments and
qualifiers during the CDL era, totaling 7,792 rounds across 852
series.\footnote{288 of the series occur in major tournaments, which are
  considered to be more ``competitive'' than the qualifiers since they
  are played in person (COVID-permitting), whereas the qualifiers are
  played online.} Data was collected in spreadsheets by community member
``IOUTurtle''.\footnote{Data: \url{https://linktr.ee/CDLArchive}.
  Author: \url{https://twitter.com/IOUTurtle}}

@@

There have been a handful of studies of the distribution of games played
in a series of a professional sport. Most assume a constant probability
\(\phi\) of a given team winning a game in the series, regardless of the
series state. Mosteller (1952) observed that the American League had
dominated the National League in Major League Baseball's (MLB) World
Series matchups, implying that matchups should not modeled with
\(\phi = 0.5\). Mosteller proposed three approaches for identifying the
optimal constant probability value of the stronger team in the World
Series, finding \(\phi \approx 0.65\). in each case: (1) solving for
\(p\) from the empirical average number of games won by the loser of the
series, which he called the ``method of moments'' approach; (2)
maximizing the likelihood that the sample would have been drawn from a
population in which the probability of a team winning a game is constant
across the series (i.e.~maximum likelihood), and (3) minimizing the
chi-square goodness of fit statistic for \(\phi\).

Chance (2020) re-examines the constant probability notion in Major
League Baseball's (MLB) World Series (1923--2018), the National
Basketball Association's (NBA) Finals (1951--2018), and the National
Hockey League's (NHL) Stanley Cup (1939--2018). Chance finds strong
evidence against the null hypothesis of \(\phi = 0.5\) in the MLB and
NHL championship series when applying Mosteller's first and second
methods. Chance goes on to outline a conditional probability framework
(likelihood of winning a game given the series state) which can exactly
explain the distribution of the number of games played.

Momentum, one of most discussed topics in sports analytics, goes
hand-in-hand with a discussion of the nature of series
outcomes.\footnote{We often use use ``streaks'' and momentum
  interchangeably, but as (Steeger, Dulin, and Gonzalez 2021) note,
  momentum implies dependence between events, whereas streaking does
  not.} Two opposing fallacies are observed in the context of momentum:
the ``gambler's fallacy'' (negative recency) and ``hot hand fallacy''
(positive recency). Per Ayton et al. (2004), negative recency is ``the
belief that, for random events, runs of a particular outcome \ldots{}
will be balanced by a tendency for the opposite outcome'', while
positive recency is the expectation of observing future results that
match recent results.

@@

Despite the plethora of existing research on games played in a series
and momentum in sports, these topics have yet to be investigated heavily
in esports. Work has been done to examine in-round win probability in
other FPS titles such as Counter-Strike (Xenopoulos, Freeman, and Silva
2022) and Valorant (DeRover 2021), both of which are round-based like
CoD SnD. However, research on round-level trends is sparse, perhaps
because games like Counter-Strike and Valorant both have economic
aspects that can create clear advantages on side in a given round, given
how prior rounds played out.

Additionally, both Counter-Strike and Valorant have overtime rules and
blocked offensive/defensive roles (i.e.~playing either offense or
defense for many consecutive rounds). On the other hand, teams in CoD
SnD rotate sides every round, analogous to a 1-1-1-1-1-1-1 format for
home advantage in best-of-seven series for professional sports like the
MLB, NBA, and NHL.\footnote{1-1-1-1-1-1-1 is not used today in these
  leagues, but it was at least one in each league.} While theoretically
one might be able to account for any kind of format, such as a 5-5-1,
the rotation of team sides every round is convenient for convincing
ourselves that rounds could reasonably be modeled as i.i.d. Bernoulli
trials.

@@

\hypertarget{sec:method-rounds-played}{%
\subsection{Distribution of rounds
played}\label{sec:method-rounds-played}}

@@

\hypertarget{sec:method-momentum}{%
\subsection{Momentum}\label{sec:method-momentum}}

@@

\begin{equation}\protect\hypertarget{eq:pwr}{}{
p_0(\text{win} | r) = p^{w|r}_0 = \{
\begin{array}{rl}
\frac{m}{r}, & \text{team wins series} \\
\frac{r - m}{r}, & \text{team loses series}
\end{array}
}
\label{eq:pwr}
\end{equation}

@@

As shown by MS with their Theorem 1, we should expect the proportion of
rounds wins immediately following a streak of \(k\) rounds wins for a
series lasting \(r\) rounds, \(\hat{P}^{w|kr}_{MS}\), to be strictly
less than \(p^{w|r}\).\footnote{See Appendix E.1 of MS for details on
  the proof that follows.} Unfortunately, there does not exist a closed
form representation for the expectation for \(\hat{P}^{w|kr}_{MS}\) for
\(k > 1\). Nonetheless, one may recursively build a collection of
dictionaries to compute the expectation analytically, or, more
conveniently, one may run simulations to estimate the expected
value.\footnote{I've adapted the code from Vafa (2017), which implements
  MS's framework (2018).}

Although it would be convenient to emulate the hypothesis testing for
difference of proportions performed by MS (and GVT), my context is
fundamentally different from that of MS, who focus on longitudinal data
in controlled settings.\footnote{GVT also performs statistical tests on
  shots from players in live games, i.e.~``observational'' data, but
  they note that their findings are likely affected player shot
  selection in the face of defensive strategy by the opposing team.} The
number of trials is fixed in their experimental designs, but in CoD SnD,
the number of rounds played is determined as a function of the max
number of possible rounds (\(s\)) and whether or not the team wins the
series, as shown in Equation \ref{eq:pwr}. The Bernoulli trial success
probability, i.e.~a round win in CoD SnD, is not independent of the
opponent. Consequently, a statistical test of the difference in
\(\hat{P}^{w|kr}\) and \(\hat{P}^{\ell|kr}\), as performed by MS to
evaluate their hypothesis regarding post-streak success rates is not
completely appropriate, although useful as a reference.

One might consider another form of the expected proportion of rounds won
immediately after a streak of \(k\) round wins in a best-of-\(s\) series
given the length of the series (\(r\) rounds), the ``notional''
proportion \(\hat{P}^{w|kr}_0\). The Bernoulli round win probability
underlying \(\hat{P}^{w|kr}_0\) is

\begin{equation}\protect\hypertarget{eq:pwkr}{}{
p_0(\text{win} | k, r) = p^{w|kr}_0 =\{
\begin{array}{rl}
\frac{m - k}{i - k}, & \text{team wins series} \\
\frac{s - m - k}{i - k}, & \text{team loses series}
\end{array}
}
\label{eq:pwkr}
\end{equation}

using \(m\) from Equation \ref{eq:m}. We can perform a binomial test to
test the null hypothesis \(H_0 : \omega = \omega_0\) for the observed
probability of success \(\omega\) and a user-specified \(\omega_0\),
where \(0 \leq \omega_0 \leq 1\).\footnote{\(\omega\) and other symbols
  are chosen so as to not conflict with symbols already used elsewhere
  for in other equations, either generic or for CoD SnD specifically.}

If there are \(r_+\) observed successes in a sample of \(r\) trials and
we expect that there should be \(r * \omega_0\), the probability of
arriving at this expected number of successes is

\begin{equation}\protect\hypertarget{eq:binom}{}{
\Pr(r_+) = {\binom {r}{r_+}} \omega^{r_+}(1-\omega)^{r-r+}.
}\label{eq:binom}\end{equation}

Plugging in the notional probability \(p^{w|kr}_0\) for the null
\(\omega_0\) and the observed probability
\(p^{w|kr} = N^{w|kr} \times P^{w|kr}\), for the observed \(\omega\),
where \(N^{w|kr}\) is the observed number of instances where a teams
wins the round immediately following a \(k\)-round winning streak in a
series lasting \(r\) rounds, we can evaluate the null hypothesis that
the observed probability is equal to the notional probability. If we
reject this null hypothesis, then we might consider team momentum,
represented by post-streak success, plausible.

One can perform the same binomial test for the MS probability adjusting
for streak-selection bias, \(p^{w|kr}_{MS}\). Given the caveats
mentioned before, the results should be heeded with caution.

Stepping back from the one-sided perspective of a single team's
probability of winning and their streakiness, one can attempt to detect
the ``hot hand'' more broadly with the Wald-Wolfowitz runs test, a
common tool used in the field of quality control. Under the null
hypothesis, the number of runs in a sequence of \(r\) trials is a random
variable with the following mean \(\mu\) and variance \(\sigma^2\).

\begin{equation}\protect\hypertarget{eq:ww}{}{
\mu = \frac{2r_+r_-}{r} + 1, \sigma^2 = \frac{(\mu-1)(\mu-2)}{r-1}
}\label{eq:ww}\end{equation}

One might subset observed series sequences to those that violate the run
test null hypothesis and perform a proportions test where the null is
that the observe rate of such series is equal to the frequency that we
should expect, given a choice for the constant round win probability
\(\phi\). If the

@@

\hypertarget{sec:results-rounds-played}{%
\subsection{Distribution of rounds
played}\label{sec:results-rounds-played}}

@@

\hypertarget{sec:results-momentum}{%
\subsection{Momentum}\label{sec:results-momentum}}

Given that people typically perceive streaks as beginning after the
third success (or failure) at minimum (Carlson and Shu 2007), I focus on
streaks of three round wins.\footnote{Three happens to also be a
  reasonable number for series that lasts at maximum 11 rounds.} Table
\ref{tbl:cod-pw3r-pl3r} compares the observed round win rate
\(P^{w,kr}\) following streaks of \(k=3\) round wins given the series
outcome (\(r\) rounds and winner) with the notional and MS expected
proportions, \(\hat{P}^{w|kr}_0\) and \(\hat{P}^{w|kr}_{MS}\)
respectively.

\begin{longtable}{rrrrrr}
\caption{Observed count and proportion of round wins immediately following round win streak of $k=3$, $N^{w|kr}$ and $P^{w|kr}$ respectively, given the length of the series ($r$ rounds) and the series winner. Additionally, the notional and MS expected proportions, $\hat{P}^{w|kr}_0$ and $\hat{P}^{w|kr}_{MS}$ respectively, and the $p$-values of binomial tests comparing the underlying notional and MS round win probabilities, $p^{w|kr}_0$ and $p^{w|kr}_{MS}$, to the probability implied by the observed proportion $p^{w|kr} = N^{w|kr} \times P^{w|kr}$.}
\label{tbl:cod-pw3r-pl3r} \\
\toprule
$r$ & \text{Win series?} & $N^{w|kr}$ & $P^{w|kr}$ & $\hat{P}^{w|kr}_0$ ($p$-value) & $\hat{P}^{w|kr}_{MS}$ ($p$-value) \\ 
\midrule
7 & yes & 209 & 74.6\% & 75.0\% (0.94) & 75.7\% (0.69) \\ 
8 & yes & 209 & 62.2\% & 60.0\% (0.53) & 61.9\% (0.94) \\ 
9 & yes & 193 & 51.8\% & 50.0\% (0.67) & 52.7\% (0.83) \\ 
10 & no & 60 & 13.3\% & 14.3\% (1.00) & 26.7\% (0.02) \\ 
10 & yes & 151 & 43.7\% & 42.9\% (0.87) & 44.7\% (0.87) \\ 
11 & no & 129 & 24.0\% & 25.0\% (0.84) & 30.8\% (0.11) \\ 
11 & yes & 150 & 40.0\% & 37.5\% (0.56) & 38.8\% (0.80) \\

@@

We observe that, with the exception of \(\hat{P}^{w|kr}_{MS}\) when
\(r = 10\) and the eventual series loser is the team that wins after a
streak of three round wins\footnote{This implies that the opponent, th
  eventual series winner, goes on to win the series 6-4 after going down
  0-4. This is the only scenario in which one team wins four rounds
  consecutively but loses a best-of-11 series in 10 rounds.}, all
binomial test \(p\)-values are greater than the typical
\(\alpha = 0.05\) threshold, implying that we cannot reject the null
hypothesis that the expected notional and MS post-streak round win rates
are different than the observed proportion. As noted before, the MS
expectations may be unreliable for CoD SnD, so one is inclined to prefer
the results of the notional binomial tests.

When performing the same tests for streaks of two, four, and five, there
is no case in which I can reject the binomial null hypothesis for the
expected notional probability \(p^{w|kr}_0\). The null hypothesis can be
rejected for expected MS probability \(p^{w|kr}_{MS}\) when
\(k = 4, r = [7, 8, 10]\) and \(k = 5, r = [7, 8]\). Nonetheless, the
usefulness of \(p^{w|kr}_{MS}\) is questionable, as described before,
and the sample sizes are smaller for these conditions (all under 20
instances, with the exception of \(k = 4, r = [7, 8]\) with 70 and 39
observations respectively).

Now let us explicitly consider the round, \(i\). As shown in Table
\ref{tbl:cod-pw3ri-pl3ri}, the notional win rate in the round \(i\)
immediately following a streak of \(k\) round wins in a series lasting
\(r\) rounds, \(\hat{P}{w|kri}_0\), is statistically different than the
observed \(P^{w|kri}\) in several cases, notably when \(r = i\) for each
of \(r \in [9, 10, 11]\).\footnote{The null can also be rejected for
  \(r \in [7, 8]\), although these are not shown, and the sample sizes
  are smaller.} Interestingly, the observed rate is greater than the
expected rate in each case where we can reject the null hypothesis,
whereas the observed rate is less than the expected rate in all other
cases (as well in a strong majority of cases not shown).

\begin{longtable}{rrrrrr}
\caption{Given the round $i$, the round win streak $k=3$, the length of the series ($r$ rounds), and the series winner, the observed count and proportion of round wins immediately following the streak, $N^{w|kri}$ and $P^{w|kri}$ respectively. Additionally, the notional proportion, $\hat{P}^{w|kri}_0$, and the $p$-value of a binomial test comparing the underlying notional $p^{w|kri}_0$ to the probability implied by the observed proportion $p^{w|kri} = N^{w|kri} \times P^{w|kri}$. Table restricted to $r \in [9, 10, 11]$ and $r - 2 \leq i \leq r$ for brevity.}
\label{tbl:cod-pw3ri-pl3ri} \\
\toprule
$r$ & $i$ & \text{Win series?} & $N^{w|kri}$ & $P^{w|kri}$ & $\hat{P}^{w|kri}_0$ ($p$-value) & $\hat{P}^{w|kri}_0 > P^{w|kri}$ \\

9 & 7 & yes & 36 & 41.7\% & 50.0\% (0.41) & no \\ 
9 & 8 & yes & 40 & 45.0\% & 50.0\% (0.64) & no \\ 
9 & 9 & yes & 37 & 89.2\% & 50.0\% (<=0.01) & yes \\ 
10 & 8 & no & 37 & 2.7\% & 14.3\% (0.06) & no \\ 
10 & 8 & yes & 37 & 24.3\% & 42.9\% (0.03) & no \\ 
10 & 9 & no & 32 & 6.2\% & 14.3\% (0.31) & no \\ 
10 & 9 & yes & 32 & 12.5\% & 42.9\% (<=0.01) & no \\ 
10 & 10 & yes & 29 & 72.4\% & 42.9\% (<=0.01) & yes \\ 
11 & 9 & no & 32 & 15.6\% & 25.0\% (0.31) & no \\ 
11 & 9 & yes & 32 & 31.2\% & 37.5\% (0.58) & no \\ 
11 & 10 & no & 36 & 13.9\% & 25.0\% (0.18) & no \\ 
11 & 10 & yes & 36 & 13.9\% & 37.5\% (<=0.01) & no \\ 
11 & 11 & yes & 32 & 53.1\% & 37.5\% (0.10) & yes \\

@@

>>>>

\begin{abstract}
My research investigates patterns in round win percentages in
professional Search and Destroy (SnD) matches of the popular
first-person shooter game Call of Duty (CoD). First, I find evidence
supprting the hypothesis that round win probability can be modeled as a
constant across the series, although not at the naive 50\%. Second, I
examine post-streak round win probability, given the series outcome. I
find no evidence that post-streak win rate is significantly different
than expected for combinations of streak length, series length, and
series winner, suggesting that the perception of momentum at the team
level is just that, perception.
\end{abstract}

@@

I adopt the terminology ``series'' to refer to what CoD SnD players
typically call a ``match'', so as to emulate the terminology of playoff
series in professional leagues like the National Basketball Association,
National Hockey League, and Major League Baseball. A ``game'' or a
``match'' in such leagues is analogous to a ``round'' of CoD SnD.
Further, it is analogous to a trial in statistical analysis of Bernoulli
events.

@@

The data set consists of all SnD matches played in major tournaments and
qualifiers during the CDL era, totaling 7,792 rounds across 852
series.\footnote{288 of the series occur in major tournaments, which are
  considered to be more ``competitive'' than the qualifiers since they
  are played in person (COVID-permitting), whereas the qualifiers are
  played online.} Data was collected in spreadsheets by community member
``IOUTurtle''.\footnote{Raw data: \url{https://linktr.ee/CDLArchive}.
  Author: \url{https://twitter.com/IOUTurtle}. Processed data can be
  found at
  \url{https://github.com/tonyelhabr/fps_round_games/blob/master/data/cod_rounds.csv}.}

@@

There have been a handful of studies of the distribution of games played
in a series of a professional sport. Most assume a constant probability
\(\phi\) of a given team winning a game in the series, regardless of the
series state. Mosteller (1952) observed that the American League had
dominated the National League in Major League Baseball's (MLB) World
Series matchups, implying that games should not be modeled with a
constant (null) \(\phi_0 = 0.5\). Mosteller proposed three approaches
for identifying the optimal constant probability value of the stronger
team in the World Series, finding \(\phi \approx 0.65\) in each case:
(1) solving for \(p\) from the empirical average number of games won by
the loser of the series, which he called the ``method of moments''
approach; (2) maximizing the likelihood that the sample would have been
drawn from a population in which the probability of a team winning a
game is constant across the series (i.e.~maximum likelihood), and (3)
minimizing the chi-square goodness of fit statistic \(\chi^2\) for
\(\phi\).

Chance (2020) re-examines the constant probability notion in Major
League Baseball's (MLB) World Series (1923--2018), the National
Basketball Association's (NBA) Finals (1951--2018), and the National
Hockey League's (NHL) Stanley Cup (1939--2018). Chance finds strong
evidence against the null hypothesis of \(\phi_0 = 0.5\) in the MLB and
NHL championship series when applying Mosteller's first and second
methods. Chance goes on to outline a conditional probability framework
(likelihood of winning a game given the series state) which can exactly
explain the distribution of the number of games played.

Momentum, one of most discussed topics in sports analytics, goes
hand-in-hand with a discussion of the nature of series
outcomes.\footnote{We often use use ``streaks'' and momentum
  interchangeably, but as (Steeger, Dulin, and Gonzalez 2021) note,
  momentum implies dependence between events, whereas streaking does
  not.} Two opposing fallacies are observed in the context of momentum:
the ``gambler's fallacy'' (negative recency) and the ``hot hand
fallacy'' (positive recency). Per Ayton et al. (2004), negative recency
is ``the belief that, for random events, runs of a particular outcome
\ldots{} will be balanced by a tendency for the opposite outcome'',
while positive recency is the expectation of observing future results
that match recent results.

@@

Other research has borrowed a common approach from the field of quality
control, looking at unlikely sequences of events with the Wald-Wolfowitz
runs test (Peel and Clauset 2015), (Steeger, Dulin, and Gonzalez 2021).
Peel and Clauset find no evidence for unlikely sequences of scoring
events in the NHL, College Football (CFB), and National Football League
(NFL), but do in the NBA. As a check on their entropy approach to
momentum identification, Steeger et al.~found three NHL teams who
sequence of wins in the 2018-2019 regular season violated the
Wald-Wolfowitz null hypothesis, although these teams did not match the
teams identified with their primary approach.

\hypertarget{my-contribution}{%
\subsection{My contribution}\label{my-contribution}}

Despite the plethora of existing research on games played in a series
and momentum in sports, these topics have yet to be investigated
heavily, if at all, in esports. Work has been done to examine
intra-round win probability in other FPS titles such as Counter-Strike
(Xenopoulos, Freeman, and Silva 2022) and Valorant (DeRover 2021), both
of which are round-based like CoD SnD. However, considering
Counter-Strike and Valorant specifically, research on round-level trends
seems non-existant, perhaps for one of many reasons:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Both have economic aspects that can create clear advantages for one
  side in a given round, given how prior rounds played out. CoD has no
  such equivalent, except for perhaps ``kill streaks'', which
  infrequently play a role.
\item
  Teams play either offense or defense for many consecutive rounds. (The
  former has a 14-14-1 format and the latter uses a 12-12-1 format.) On
  the other hand, teams in CoD SnD rotate roles every round, analogous
  to a 1-1-1-1-1-1-1 format for home advantage in best-of-seven series
  for professional sports like the MLB, NBA, and NHL.\footnote{1-1-1-1-1-1-1
    is not used today in these leagues, but it was at least one in each
    league.} While theoretically one might be able to account for any
  kind of format, such as a 5-5-1, the rotation of team sides every
  round is convenient for convincing ourselves that rounds could
  reasonably be modeled as i.i.d. Bernoulli trials.
\item
  Both Counter-Strike and Valorant have overtime rules---team must win
  by two rounds---which can make end-of-series sequences difficult to
  model. CoD SnD does not have overtime rules.
\end{enumerate}

To my knowledge, there is no existing public statistical research on CoD
SnD format, beyond descriptive analysis on social media. While
intra-round trends may be more directly applicable to teams looking for
an advantage on their competition, broader investigation on a common
concept like the hot-hand phenomenon in a new field---particularly one
that is less subjected to factors that may be difficult to control for,
e.g.~weather---should hopefully be useful as a precedent or supplement
for future researchers.

@@

\hypertarget{distribution-of-rounds-played}{%
\subsection{Distribution of rounds
played}\label{distribution-of-rounds-played}}

@@

\hypertarget{momentum}{%
\subsection{Momentum}\label{momentum}}

\hypertarget{ms-post-streak-probability}{%
\subsubsection{MS post-streak
probability}\label{ms-post-streak-probability}}

@@

\begin{equation}\protect\hypertarget{eq:pwr}{}{
p_0(\text{win} | r) = p^{+|r}_0 = \{
\begin{array}{cl}
\frac{m}{r}, & \text{team wins series} \\
\frac{r - m}{r}, & \text{team loses series}
\end{array}
}
\label{eq:pwr}
\end{equation}

@@

As shown by MS with their Theorem 1, we should expect the proportion of
rounds wins immediately following a streak of \(k\) rounds wins for a
series lasting \(r\) rounds, \(\hat{P}^{+|kr}_{MS}\), to be strictly
less than \(p^{+|r}\).\footnote{See Appendix E.1 of MS for the proof.}
Unfortunately, there does not exist a closed form representation for the
expected value of \(\hat{P}^{+|kr}_{MS}\) for \(k > 1\). Nonetheless,
one may recursively build a collection of dictionaries to compute the
expectation analytically, or, more conveniently, one may run simulations
to estimate the expected value, as I choose to do.\footnote{I have
  adapted the code from Vafa (2017), which implements MS's framework
  (2018).}

Although it would be convenient to emulate the hypothesis testing for
difference of proportions performed by MS (and GVT), my context is
fundamentally different from that of MS, who focus on longitudinal data
in controlled settings.\footnote{GVT also performs statistical tests on
  shots from players in live games, i.e.~``observational'' data, but
  they note that their findings are likely affected player shot
  selection in the face of defensive strategy by the opposing team.} The
number of trials is fixed in their experimental designs, but in CoD SnD,
the number of rounds played is determined as a function of the max
number of possible rounds (\(s\)) and whether or not the team wins the
series. The Bernoulli trial success probability, i.e.~a round win in CoD
SnD, is not independent of the opponent. Consequently, a statistical
test of the difference in \(\hat{P}^{+|kr}\) and \(\hat{P}^{-|kr}\), as
performed by MS to evaluate their hypothesis regarding post-streak
success rates, is not completely appropriate, although useful as a
reference.

\hypertarget{notional-post-streak-probability}{%
\subsubsection{``Notional'' post-streak
probability}\label{notional-post-streak-probability}}

One might consider another form of the expected proportion of rounds won
immediately after a streak of \(k\) round wins in a best-of-\(s\) series
given the length of the series (\(r\) rounds), the ``notional''
proportion \(\hat{P}^{+|kr}_0\). The Bernoulli round win probability
underlying \(\hat{P}^{+|kr}_0\) is

\begin{equation}\protect\hypertarget{eq:pwkr}{}{
p_0(\text{win} | k, r) = p^{+|kr}_0 =\{
\begin{array}{cl}
\frac{m - k}{r - k}, & \text{team wins series} \\
\frac{s - m - k}{r - k}, & \text{team loses series}
\end{array}
}
\label{eq:pwkr}
\end{equation}

We can perform a binomial test to evaluate the null hypothesis
\(\omega = \omega_0\) for the observed probability of success \(\omega\)
and a user-specified \(\omega_0\), where
\(0 \leq \omega_0 \leq 1\).\footnote{\(\omega\) and other symbols are
  chosen so as to not conflict with symbols already used elsewhere for
  in other equations, either generic or for CoD SnD specifically.} If
there are \(r^+\) observed successes in a sample of \(r\) trials and we
expect that there should be \(r * \omega_0\), the probability of
arriving at this expected number of successes is

\begin{equation}\protect\hypertarget{eq:binom}{}{
\Pr(r^w) = {\binom {r}{r^+}} \omega^{r^+}(1-\omega)^{r-r^+}.
}\label{eq:binom}\end{equation}

Plugging in the notional proportion \(\hat{p}^{+|kr}_0\) for the null
\(\omega_0\) and the observed proportion \(P^{+|kr}\) for \(\omega\)
(treating proportions as probabilities), we can evaluate the null
hypothesis that the observed probability is equal to the notional
probability. If we can reject this null hypothesis, then we might
consider team momentum, represented by post-streak success, plausible.

One can perform the same binomial test for the MS's
streak-selection-adjusted proportion, \(\hat{P}^{+|kr}_{MS}\). FOOTY:
Given the caveats mentioned before, the results should be heeded with
caution.

We can further decompose Equation \ref{eq:pwkr} by the round \(i\)
(\(i \leq r\)) in which the streak of length \(k\) carries into.

\begin{equation}\protect\hypertarget{eq:pwkri}{}{
p_0(\text{win} | k, r, i) = p^{+|kri}_0 =\{
\begin{array}{cl}
\frac{m - k}{r - k}, & \text{team wins series}, i \neq r \\
1, & \text{team wins series}, i = r \\
\frac{s - m - k}{r - k}, & \text{team loses series}, i \neq r \\
0, & \text{team loses series}, i = r
\end{array}
}
\label{eq:pwkri}
\end{equation}

Again, one might apply a binomial test to evaluate the hypothesis that
the expected proportions, both \(\hat{P}^{+|kri}_0\) and
\(\hat{P}^{+|kri}_{MS}\), are equal to the observed proportion
\(P^{+|kr}\).

\hypertarget{wald-wolfowitz-runs-test}{%
\subsubsection{Wald-Wolfowitz runs
test}\label{wald-wolfowitz-runs-test}}

Stepping back from the one-sided perspective of a single team's round
win probability when streaking, one can attempt to detect the hot hand
phenomenon with the Wald-Wolfowitz runs test. Under the null hypothesis,
the number of runs in a sequence of \(r\) trials is a random variable,
which can take on values \(r^+\) and \(r^-\) for success and failure
respectively, with the following mean \(\mu\) and variance \(\sigma^2\).

\begin{equation}\protect\hypertarget{eq:ww}{}{
\mu = \frac{2r^{+}r^{-}}{r} + 1, \sigma^2 = \frac{(\mu-1)(\mu-2)}{r-1}
}\label{eq:ww}\end{equation}

One might subset observed series sequences to those that violate the run
test null hypothesis and perform a test of proportions, where the null
is that the observed rate of such series is equal to the frequency that
we should expect, given a choice for the constant round win probability
\(\phi\). If we can reject the null, then we can argue in favor of the
hot hand.

@@

\hypertarget{distribution-of-rounds-played-1}{%
\subsection{Distribution of rounds
played}\label{distribution-of-rounds-played-1}}

@@

\hypertarget{momentum-1}{%
\subsection{Momentum}\label{momentum-1}}

Given that people typically perceive streaks as beginning after the
third success (or failure) at minimum (Carlson and Shu 2007), I focus on
streaks of three round wins.\footnote{Three happens to also be a
  reasonable number for series that lasts at maximum 11 rounds.} Table
\ref{tbl:cod-pw3r-pl3r} compares the observed round win rate
\(P^{w,kr}\) following streaks of \(k=3\) round wins given the series
outcome (\(r\) rounds and winner) with the notional and MS expected
proportions, \(\hat{P}^{+|kr}_0\) and \(\hat{P}^{+|kr}_{MS}\)
respectively.

\begin{longtable}{rcrrrrr}
\caption{Given the round win streak $k$, the length of the series ($r$ rounds), and the series winner, the observed count of rounds wins, $r^{+|kr}$, and proportion of round wins, $P^{+|kr}$, among $N^{kr}$ instances where a team could win after the streak $k=3$ ($P^{+|kr} = \frac{r^{+|kr}}{N^{kr}}$). Additionally, the notional and MS expected proportions, $\hat{P}^{+|kr}_0$ and $\hat{P}^{+|kr}_{MS}$ respectively.}
\label{tbl:cod-pw3r-pl3r} \\

\toprule
$r$ & \text{Win series?} & $r^{+|kr}$ & $N^{kr}$ & $P^{+|kr}$ & $\hat{P}^{+|kr}_0$ & $\hat{P}^{+|kr}_{MS}$ \\ 
\midrule

7 & yes & 156 & 209 & 74.6\% & 75.0\% & 75.7\% \\ 
8 & yes & 130 & 209 & 62.2\% & 60.0\% & 61.9\% \\ 
9 & yes & 100 & 193 & 51.8\% & 50.0\% & 52.7\% \\ 
10 & no & 8 & 60 & 13.3\% & 14.3\% & 26.7\% \\ 
10 & yes & 66 & 151 & 43.7\% & 42.9\% & 44.7\% \\ 
11 & no & 31 & 129 & 24.0\% & 25.0\% & 30.8\% \\ 
11 & yes & 60 & 150 & 40.0\% & 37.5\% & 38.8\% \\

@@

With the exception of \(\hat{P}^{+|kr}_{MS}\) when \(r = 10\) and the
eventual series loser is the team that wins after a streak of three
round wins\footnote{This implies that the opponent, the eventual series
  winner, goes on to win the series 6-4 after going down 0-4. This is
  the only scenario in which one team wins four rounds consecutively but
  loses a best-of-11 series in 10 rounds.}, all un-adjusted binomial
test \(p\)-values are greater than the typical \(\alpha = 0.05\)
threshold, implying that we cannot reject the null hypothesis that the
expected notional and MS post-streak round win rates are different than
the observed proportion. Adjusting the \(p\)-values with the Benjamini
\& Yekutieli (BY) correction (2001), the null hypothesis cannot be
rejected for any case.\footnote{As noted before, the MS expectations may
  be unreliable for CoD SnD, so one is inclined to prefer the results of
  the notional binomial tests.}

When performing the same tests for streaks of two, four, and five, there
is no case in which I can reject the binomial null hypothesis for the
expected notional probability \(p^{+|kr}_0\). The null hypothesis can
only be rejected for the expected MS probability \(p^{+|kr}_{MS}\) when
\(k = 5, r = 7\) after applying the BY p-value correction.\footnote{At
  36, the sample size for \$k = 5, r = 7\$ is not as large as most other
  combinations of \$k\$ and \$r\$.}

Now let us explicitly consider the round, \(i\). As shown in Table
\ref{tbl:cod-pw3ri-pl3ri}, the notional win rate in the round \(i\)
immediately following a streak of \(k\) round wins in a series lasting
\(r\) rounds, \(\hat{P}{+|kri}_0\), is statistically different than the
observed \(P^{+|kri}\) in several cases, notably when \(r = i\) for each
of \(r \in [9, 10, 11]\).\footnote{The null can also be rejected for
  \(r \in [7, 8]\), although these are not shown, and the sample sizes
  are smaller.} Interestingly, the observed rate is greater than the
expected rate in each case where we can reject the null hypothesis,
whereas the observed rate is less than the expected rate in all other
cases, as well in a strong majority of cases not shown.

\begin{longtable}{rrcrrrr}
\caption{Given the round $i$, the round win streak $k$, the length of the series ($r$ rounds), and the series winner, the observed count of rounds wins, $r^{+|kri}$, and proportion of round wins, $P^{+|kri}$, among $N^{kri}$ instances where a team could win after the streak $k=3$ ($P^{+|kri} = \frac{r^{+|kri}}{N^{kri}}$). Additionally, the notional proportion, $\hat{P}^{+|kri}_0$, and the $p$-value of a binomial test comparing $\hat{P}^{+|kri}_0$ to the observed proportion $P^{+|kri}$. Table restricted to $r \in [10, 11]$ and $i < r$ and $N^{kri} > 10$ for brevity.}
\label{tbl:cod-pw3ri-pl3ri} \\
\toprule
$r$ & $i$ & \text{Win series?} & $r^{+|kri}$ & $N^{kri}$ & $P^{+|kri}$ & $\hat{P}^{+|kri}_0$\\ 
\midrule

10 & 7 & no & 2 & 13 & 15.4\% & 14.3\% \\ 
10 & 7 & yes & 11 & 13 & 84.6\% & 42.9\% \\ 
10 & 8 & no & 1 & 10 & 10.0\% & 14.3\% \\ 
10 & 8 & yes & 9 & 10 & 90.0\% & 42.9\% \\ 
11 & 5 & no & 9 & 13 & 69.2\% & 25.0\% \\ 
11 & 5 & yes & 4 & 13 & 30.8\% & 37.5\% \\ 
11 & 6 & no & 4 & 13 & 30.8\% & 25.0\% \\ 
11 & 6 & yes & 9 & 13 & 69.2\% & 37.5\% \\ 
11 & 7 & no & 3 & 10 & 30.0\% & 25.0\% \\ 
11 & 7 & yes & 7 & 10 & 70.0\% & 37.5\% \\ 
11 & 9 & no & 5 & 15 & 33.3\% & 25.0\% \\ 
11 & 9 & yes & 10 & 15 & 66.7\% & 37.5\% \\ 
11 & 10 & no & 5 & 10 & 50.0\% & 25.0\% \\ 
11 & 10 & yes & 5 & 10 & 50.0\% & 37.5\% \\

@@

In Table \ref{tbl:cod-runs}, the observed and expected count of
sequences, \(N^{\zeta r}\) and \(\hat{N}^{\zeta r}\) respectively, are
shown for sequences, \(\zeta(r)\), for which one can reject the
Wald-Wolfowitz null hypothesis at a confidence level of
\(\alpha = 0.05\), for \(r \in [8, 9, 10]\). (In addition to these 13,
there are 12 additional sequences for \(r = 11\).) The relative
frequency of the expected proportions are based on 10,000 simulations
using the constant round probability \(\phi_2 = 0.575\). A test for
difference between the actual and expected proportions of all sequences
indicates that the null hypothesis---that the two proportions are
equal---cannot be rejected for any of the 25 significant sequences.

\begin{longtable}{rlrrrr}
